Bayesian Hyper-Heuristics

### 2023-01-23 article/sections/05-probability.tex
🚧  (article): Probability WIP

\chapter{Probability}
\label{chap:probability}

\begin{quotation}
      ``Probability theory is nothing but common sense reduced to calculation.''
\end{quotation}
\begin{flushright}
      - Pierre-Simon Laplace
\end{flushright}

@@

Chapter \ref{chap:hhs} provided the concept of a \acf{HH}. This dissertation aims to develop a selection \acs{HH} that makes use of probability theory to select the best \acs{HH} to solve a given problem. This chapter aims to provide the necessary background information on probability theory and statistics. These are large fields and focus is put on the elements that are required to formulate the proposed \acf{BHH}. The remainder of the chapter is structured as follows:

\begin{itemize}
      \item \textbf{Section \ref{sec:probability:overview}} provides a brief overview of what probability is and how it is used.

      \item \textbf{Section \ref{sec:probability:cond_probability}} presents the concept of conditional probability.

      \item \textbf{Section \ref{sec:probability:multiple_events}} presents the two laws of probability related to the intersection and union of multiple events.

      \item \textbf{Section \ref{sec:probability:bayes_theorem}} introduces \index{Bayes' theorem}Bayes' theorem, the fundamental theorem upon which the \acs{BHH} is built.

      \item \textbf{Section \ref{sec:probability:probability_distributions}} presents the reader with relevant probability distributions.

      \item \textbf{Section \ref{sec:probability:conjugate_priors}} presents the reader with relevant conjugate prior probability distributions.

      \item \textbf{Section \ref{sec:probability:bayesian_statistics}} presents \textit{Bayesian} statistics. Brief discussions follow on the \textit{frequentist} view and \textit{Bayesian} view of probability. Detailed discussions follow on Bayesian optimisation methods such as \index{Bayesian analysis}\textit{Bayesian analysis}.

      \item \textbf{Section \ref{sec:probability:summary}} provides a brief summary of the chapter.
\end{itemize}

@@

\section{Conditional Probability and Independence}\label{sec:probability:cond_probability}

The occurrence of a given random event, $A$, can often be conditional on the occurrence of another event, $B$. In the field of medicine, an example of this is to calculate the probability of a certain diagnosis of a sick patient given his/her symptoms. The aforementioned relationship between events is referred to as the conditional probability between two events. The conditional probability is expressed as $P(A \vert B)$ and is read \textit{the probability of $A$ given $B$}. On the contrary, the \textit{unconditional} probability is the probability of an event, not dependent on any other. The conditional probability of $A$ given $B$ can be expressed as is given in Definition \ref{eq:probability:cond_probability} below~\cite{ref:wackerly:2014}.
\\
\begin{definition}[\textbf{Conditional Probability}]
      \label{eq:probability:cond_probability}
      The conditional probability of an event $A$, subject to the occurrence of event $B$ is expressed as

      \begin{equation}
            \label{eq:probability:overview:conditional}
            P(A \vert B) = \frac{P(A \cap B)}{P(B)}
      \end{equation}
\end{definition}

where $P(B) > 0$. Note that conditional probability does not suggest causation. If an event $A$ has a high probability of occurring after observing event $B$, it does not necessarily mean that $A$ is caused by $B$. Conditional probability simply expresses the dependence amongst events.

It could also be the case that the outcome of observing event $A$ is not affected by the occurrence of $B$. In this case, it is said that events $A$ and $B$ are independent. The independence between two events are expressed in Definition \ref{def:probability:cond_probability:independence} below.
\\
\begin{definition}[\textbf{Independence of Events}]
      \label{def:probability:cond_probability:independence}
      Two events, $A$ and $B$, are said to be independent of each other if, and only if the following criteria hold:

      \begin{itemize}
            \item $P(A \vert B) = P(A)$
            \item $P(B \vert A) = P(B)$
            \item $P(A \cap B) = P(A)P(B)$
      \end{itemize}

      Otherwise, events $A$ and $B$ are said to be \textit{dependent} random events.
\end{definition}

\section{Two Laws of Probability for Multiple Events}\label{sec:probability:multiple_events}

Suppose there are two random events, $A$ and $B$, then one can calculate the probability of the union and intersection of these events. From the aforementioned concept, two laws of probability can be formulated. These are referred to as the \textit{multiplicative} and \textit{additive} laws of probability~\cite{ref:wackerly:2014} and is given below in Theorems \ref{th:probability:multiple_events:multiplicative} and \ref{th:probability:multiple_events:additive} respectively.\\

\begin{theorem}[\textbf{The Multiplicative Law of Probability}]
      \label{th:probability:multiple_events:multiplicative}
      The probability of the intersection of two events, $A$ and $B$, is given as

      \begin{equation}
            \begin{split}
                  P(A \cap B)
                  &= P(A)P(B \vert A) \\
                  &= P(B)P(A \vert B)
            \end{split}
      \end{equation}

      If $A$ and $B$ are independent, then

      \begin{equation}
            P(A \cap B) = P(A)P(B)
      \end{equation}
\end{theorem}

\begin{proof}
      Proof is given from Definition \ref{eq:probability:cond_probability}.
\end{proof}
\vspace*{0.5cm}

\begin{theorem}[\textbf{The Additive Law of Probability}]
      \label{th:probability:multiple_events:additive}
      The probability of the union of two events, $A$ and $B$, is given as

      \begin{equation}
            P(A \cup B) = P(A) + P(B) - P(B \cap A)
      \end{equation}
\end{theorem}

\begin{proof}
      The geometric proof of the additive law of probability is given by the Venn Diagram presented in Figure \ref{fig:probability:multiple_events:additive}. Note that $A \cup B = A \cup ( \bar{A} \cap B)$, where $A$ and $\bar{A} \cap B$ are mutually exclusive events. Furthermore, consider that $B = (\bar{A} \cap B) \cup (A \cap B)$, where $\bar{A} \cap B$ and $A \cap B$ are mutually exclusive. Then $P(A \cup B) = P(A) + P(\bar{A} \cap B)$ and $P(B) = P(\bar{A} \cap B) + P(A \cap B)$. The equality on the right implies that $P(\bar{A} \cap B) = P(B) - P(A \cap B)$. By substituting the expression for $P(\bar{A} \cap B)$ into the expression for $P(A \cup B)$, the resulting expression $P(A \cup B) = P(A) + P(B) - P(A \cap B)$ is obtained.
\end{proof}

\begin{figure}[htbp]
      \centering
      \includegraphics[width=0.7\textwidth]{images/additive_law_of_probability_proof.pdf}
      \caption{A Venn-Diagram showing the proof of the additive law of probability for multiple events.}
      \label{fig:probability:multiple_events:additive}
\end{figure}

\index{Bayes' theorem}
\section{Bayes' Theorem}\label{sec:probability:bayes_theorem}

\index{Bayes' theorem}Bayes' theorem, named after Thomas Bayes, describes the probability of an event, $A$, based on prior knowledge of conditions that might be related to $A$~\cite{ref:zalta:2015}. In order to derive the formal theorem and proof, first consider Definition \ref{def:probability:bayes_theorem:partition} below~\cite{ref:zalta:2015}.
\\
\begin{definition}
      \label{def:probability:bayes_theorem:partition}
      For some positive integer, $K$, let the sets $B_{1}, B_{2}, \dots, B_{K}$ be such that

      \begin{enumerate}
            \item $S = B_{1} \cup B_{2} \cup \dots \cup B_{K}$
            \item $B_{i} \cap B_{j} = \emptyset$, for $i \neq j$
      \end{enumerate}

      Then the collection of sets $\{B_{1}$, $B_{2}$, $\dots$, $B_{K}\}$ is said to be a partition of S, the union of mutually exclusive subsets.
\end{definition}
\vspace*{0.5cm}

@@

\section{Probability Distributions}\label{sec:probability:probability_distributions}

Probability distributions are mathematical functions that give the probabilities of the occurrences of different possible outcomes in the experiment. The theory and equations presented in the following sections where all taken from~\citeauthor{ref:wackerly:2014}~\cite{ref:wackerly:2014}.

\subsection{Beta Probability Distribution}\label{sec:probability:probability_distributions:beta}

The \index{Beta probability distribution}Beta probability distribution is a family of univariate, continuous, probability distributions over some $x$, with support on the interval $[0,1]$~\cite{ref:wackerly:2014}. It is parameterised by two shape parameters $\alpha > 0$, $\alpha \in \mathbb{R}$ and $\beta > 0$, $\beta \in \mathbb{R}$. The \index{Beta probability distribution}Beta probability distribution is denoted as $Beta(\alpha, \beta)$. The \acf{PDF} of the \index{Beta probability distribution}Beta probability distribution is given as

\begin{equation}
      \label{eq:probability:probability_distributions:beta:pdf}
      P(x \vert \alpha, \beta) = f_{Beta}(x; \alpha, \beta) = \frac{1}{B(\alpha, \beta)} x^{\alpha - 1} (1 - x)^{\beta - 1}
\end{equation}

The normalising constant, $B(\alpha, \beta)$, is defined as

\begin{equation}
      \label{eq:probability:probability_distributions:beta:norm_const}
      B(\alpha, \beta) = \frac{\Gamma(\alpha)\Gamma(\beta)}{\Gamma(\alpha + \beta)}
\end{equation}

where $\Gamma$ is the Gamma function defined as

\begin{equation}
      \label{eq:probability:probability_distributions:beta:gamma_func}
      \Gamma(n) = ( n - 1)!
\end{equation}

It should be noted that the Gamma function can also be written as

\begin{equation}
      \label{eq:probability:probability_distributions:beta:gamma_func_alt}
      \Gamma(n+1) = n!
\end{equation}

The control parameters $\alpha$ and $\beta$ determine the shape of the distribution. There exists a special case where $\alpha = \beta$. This is referred to as the \textit{symmetric} \index{Beta probability distribution}Beta probability distribution. In the case where $\alpha = \beta = 1$ the distribution is equivalent to the uniform distribution over all points in its support. The \index{Beta probability distribution}Beta probability distribution for various values of $\alpha$ and $\beta$, including the symmetric version is illustrated in Figure \ref{fig:probability:probability_distributions:beta}.

\begin{figure}[htbp]
      \begin{subfigure}{0.49\textwidth}
            \centering
            \includegraphics[width=\textwidth]{images/beta_various.pdf}
            \caption{\index{Beta probability distribution}Beta probability distribution}
            \label{fig:probability:probability_distributions:beta_normal}
      \end{subfigure}
      \begin{subfigure}{0.49\textwidth}
            \centering
            \includegraphics[width=\textwidth]{images/beta_cumulative.pdf}
            \caption{Cumulative \index{Beta probability distribution}Beta probability distribution}
            \label{fig:probability:probability_distributions:beta_cumulative}
      \end{subfigure}
      \par\bigskip
      \caption{An illustration of the \index{Beta probability distribution}Beta probability distribution (left)~\cite{ref:beta:2014} as well as the cumulative \index{Beta probability distribution}Beta probability distribution (right)~\cite{ref:cumulativebeta:2014} for various values of $\alpha$ and $\beta$ .}
      \label{fig:probability:probability_distributions:beta}
\end{figure}

The expected value of $x$ is given as

\begin{equation}
      \label{eq:probability:probability_distributions:beta:expected_value}
      E[x] = \frac{\alpha}{\alpha + \beta}
\end{equation}

Similarly, the expected value of the natural logarithm of $x$ is calculated as

\begin{equation}
      \label{eq:probability:probability_distributions:beta:expected_value_ln}
      E[\ln(x)] = \psi({\alpha}) - \psi(\alpha + \beta)
\end{equation}

where $\psi$ is the logarithmic derivative of the Gamma function, called the \textit{Digamma} function. The Digamma function is given as

\begin{equation}
      \label{eq:probability:probability_distributions:beta:digamma}
      \psi(n) = \frac{d}{dn}\ln(\Gamma(n)) = \frac{\Gamma'(n)}{\Gamma(n)}
\end{equation}

Finally, the mode of the distribution is given as

\begin{equation}
      \label{eq:probability:probability_distributions:beta:mode}
      M[x] = E[x] - 1 = \frac{\alpha - 1}{\alpha + \beta - 1}
\end{equation}

\subsection{Dirichlet Probability Distribution}\label{sec:probability:probability_distributions:dirichlet}

The \index{Dirichlet probability distribution}Dirichlet probability distribution is a family of multivariate continuous probability distributions over some $\boldsymbol{x}$ in $K$ dimensions~\cite{ref:wackerly:2014}. The \index{Dirichlet probability distribution}Dirichlet probability distribution is a multivariate generalization of the \index{Beta probability distribution}Beta probability distribution and is thus sometimes referred to by its alternative name, i.e. the multivariate Beta probability distribution. The \index{Dirichlet probability distribution}Dirichlet probability distribution is parameterised by some vector $\boldsymbol{\alpha} = (\alpha_{1}$,  $\dots, \alpha_{K}), \forall_{k=1}^{K} \alpha_{k} > 0, \alpha_{k} \in \mathbb{R}$. The parameters, $\boldsymbol{\alpha}$, are referred to as the concentration parameters. The \index{Dirichlet probability distribution}Dirichlet probability distribution of order $K \geq 2$ with parameters $\boldsymbol{\alpha}$, denoted $Dir(\boldsymbol{\alpha})$, has a \acs{PDF} given as

\begin{equation}
      \label{eq:probability:probability_distributions:dirichlet:pdf}
      P(\boldsymbol{x} \vert \boldsymbol{\alpha}) =  f_{Dir}(\boldsymbol{x}; K, \boldsymbol{\alpha}) = \frac{1}{B(\boldsymbol{\alpha})}  \prod_{k=1}^{K} x_{k}^{\alpha_{k} - 1}
\end{equation}

The normalising constant, $B(\boldsymbol{\alpha})$, is defined as

\begin{equation}
      \label{eq:probability:probability_distributions:dirichlet:norm_cost}
      B(\boldsymbol{\alpha}) = \frac{\prod_{k=1}^{K} \Gamma(\alpha_{k})}{\Gamma(\alpha_{0})}
\end{equation}

where $\alpha_{0}$ is defined as

\begin{equation}
      \label{eq:probability:probability_distributions:dirichlet:alpha_0}
      \alpha_{0} = \sum_{k=1}^{K}\alpha_{k}
\end{equation}

Importantly, the set $\{x_{k}\}_{k=1}^{K}$ belongs to the standard $K-1$ probability simplex $S$, meaning that $x_{K} = 1 - \sum_{k=1}^{K-1}x_{k}$ with support $\forall_{k=1}^{K} x_{k} \in [0,1]$. Under the simplex $S$, this means that the sum over all values of the vector $\boldsymbol{x}$ must be 1. The simplex can thus be rewritten as $\sum_{k=1}^{K}x_{k} = 1$.

Similar to the \index{Beta probability distribution}Beta probability distribution, $\boldsymbol{\alpha}$ determines the shape of the distribution in $K$ dimensions and thus, there also exist a special case, referred to as the \textit{symmetric} distribution when $\forall_{k=1}^{K} \alpha_{k} = c$, where $c$ is some constant. In the case where $c = 1$, the distribution is referred to as a \textit{flat} distribution and yields the uniform distribution over all points in $S$. The \index{Dirichlet probability distribution}Dirichlet probability distribution of order $K = 3$ for various values of $\boldsymbol{\alpha}$, including the symmetric version, is presented in Figure \ref{fig:probability:probability_distributions:dirichlet}.

\begin{figure}[htb]
      \centering
      \includegraphics[width=0.9\textwidth]{images/dirichlet.pdf}
      \caption{The \acfp{PDF} for the \index{Dirichlet probability distribution}Dirichlet probability distribution over the 2-simplex. The concentration parameters, $\boldsymbol{\alpha}$, are varied. The values of the \acs{PDF} are shown by the colour maps with contour lines at equal values as indicated in the colour bars~\cite{ref:dirichlet:2020}.}
      \label{fig:probability:probability_distributions:dirichlet}
\end{figure}

The expected value of $x_{k}$ is calculated as

\begin{equation}
      \label{eq:probability:probability_distributions:dirichlet:expected_value}
      E[x_{k}] = \frac{\alpha_{k}}{\alpha_{0}}
\end{equation}

Similarly, the expected value of the natural logarithm of $x_{k}$ is calculated as

\begin{equation}
      \label{eq:probability:probability_distributions:dirichlet:expected_value_ln}
      E[\ln(x_{k})] = \psi({\alpha_{k}}) - \psi(\alpha_{0})
\end{equation}

where $\psi$ is the Digamma function as defined in Equation~\eqref{eq:probability:probability_distributions:beta:digamma}. Finally, the mode of the distribution is given as

\begin{equation}
      \label{eq:probability:probability_distributions:dirichlet:mode}
      \begin{split}
            M[x_{k}] &= E[{x_{k}] -K^{-1}} \\
            &=  \frac{\alpha_{k} - 1}{\alpha_{0} - K}
      \end{split}
\end{equation}

\subsection{Bernoulli Probability Distribution}\label{sec:probability:probability_distributions:bernoulli}

The \index{Bernoulli probability distribution}Bernoulli probability distribution is a discrete probability distribution over some random variable $x$ that takes the value of $1$ with probability $\theta$ and $0$ with probability $1-\theta$ ~\cite{ref:wackerly:2014}. The \index{Bernoulli probability distribution}Bernoulli probability distribution is denoted as $Ber(\theta)$ with support $x \in \{0, 1\}$. In probability theory, the \index{Bernoulli probability distribution}Bernoulli probability distribution is often used to explain the possible outcomes of a single experiment that asks a \textit{yes-no} question such as flipping a coin. The outcome of such an experiment is a Boolean value. The \index{Bernoulli probability distribution}Bernoulli probability distribution has a \acs{PMF}, given as

\begin{equation}
      \label{eq:probability:probability_distributions:bernoulli:pmf}
      P(x \vert \theta) = f_{Ber}(x; \theta) =
      \begin{cases}
            \theta     & \text{if}\ x=1 \\
            1 - \theta & \text{if}\ x=0
      \end{cases}
\end{equation}

Equation~\eqref{eq:probability:probability_distributions:bernoulli:pmf} can also be expressed as

\begin{equation}
      \label{eq:probability:probability_distributions:bernoulli:alt}
      f_{Ber}(x; \theta) = \theta^{x}(1-\theta)^{1-x}
\end{equation}

The mean of the \index{Bernoulli probability distribution}Bernoulli probability distribution approaches $\theta$ over many samples according to the \acs{CLT}~\cite{ref:grinstead:1997}. Figure \ref{fig:probability:probability_distributions:bernoulli:coin} illustrates a fair-coin flipping simulation. The mean converges to 0.5 for various sample sizes.

\begin{figure}[htbp]
      \centering
      \begin{subfigure}{0.49\textwidth}
            \centering
            \includegraphics[width=\textwidth]{images/coin_flip_samples_100.pdf}
            \caption{100 Samples}
            \label{fig:probability:probability_distributions:bernoulli:coin_100}
      \end{subfigure}
      \begin{subfigure}{0.49\textwidth}
            \centering
            \includegraphics[width=\textwidth]{images/coin_flip_samples_10000.pdf}
            \caption{10000 Samples}
            \label{fig:probability:probability_distributions:bernoulli:coin_10000}
      \end{subfigure}
      \par\bigskip
      \caption{An illustration of the coin-flip simulation for different sample sizes that show the convergence of the mean as per the \acf{CLT}.}
      \label{fig:probability:probability_distributions:bernoulli:coin}
\end{figure}

The expected value of the distribution is thus given as

\begin{equation}
      \label{eq:probability:probability_distributions:bernoulli:expected_value}
      E[x] = \theta
\end{equation}

The mode of the distribution is given as

\begin{equation}
      \label{eq:probability:probability_distributions:bernoulli:mode}
      M[x] =
      \begin{cases}
            0   & \text{if}\ \theta < 0.5 \\
            0,1 & \text{if}\ \theta = 0.5 \\
            1   & \text{if}\ \theta > 0.5 \\
      \end{cases}
\end{equation}

\subsection{Binomial Probability Distribution}\label{sec:probability:probability_distributions:bin}

The \index{Binomial probability distribution}Binomial probability distribution is a discrete probability distribution over a random variable $x$ taking on a number of successes in $N$ sequential independent experiments that each ask a \textit{yes-no} question~\cite{ref:wackerly:2014}. The probability of a single independent experiment yielding a success is given as $\theta$ and the \index{Binomial probability distribution}Binomial probability distribution is denoted as $Bin(N, \theta)$, with support $x \in \{0, 1, \dots, N\}$.  It should be noted that the \index{Binomial probability distribution}Binomial probability distribution is an extension of the \index{Bernoulli probability distribution}Bernoulli probability distribution over $N$ independent sequential experiments, and thus each experiment also yields some Boolean outcome. When $N=1$, the experiment is referred to as a Bernoulli trial, and the distribution is just a \index{Bernoulli probability distribution}Bernoulli probability distribution. When $N > 1$, the sequence of outcomes is referred to as a Bernoulli process. The \acs{PMF} of the \index{Binomial probability distribution}Binomial probability distribution is given as

\begin{equation}
      \label{eq:probability:probability_distributions:binomial:pmf}
      P(x \vert \theta; N) = f_{Bin}(x; N, \theta) = \binom{N}{x} \theta^{x}(1-\theta)^{1-x}
\end{equation}

Similar to the \index{Bernoulli probability distribution}Bernoulli probability distribution, the mean of the \index{Binomial probability distribution}Binomial probability distribution is $N\theta$ given the \acs{CLT} as was shown in Figure \ref{fig:probability:probability_distributions:bernoulli:coin}. The expected value of the \index{Binomial probability distribution}Binomial probability distribution is thus given as

\begin{equation}
      \label{eq:probability:probability_distributions:binomial:expected_value}
      E[x] = N\theta
\end{equation}

The mode of the distribution is given as

\begin{align}
      \label{eq:probability:probability_distributions:binomial:mode}
      \begin{split}
            M[x] &= E[x] + \theta \\
            &= N\theta  + \theta \\
            &= (N  + 1)\theta
      \end{split}
\end{align}

\subsection{Categorical Probability Distribution}\label{sec:probability:probability_distributions:categorical}

The \index{Categorical probability distribution}Categorical probability distribution is a discrete probability distribution over some random variable, $x$, taking on any one of $K$ possible categories~\cite{ref:wackerly:2014}. There is no ordering to these categories and therefore, for simplicity, each category is assigned a numerical representative value such that $k \in \{1, 2, \dots, K\}$. The probabilities for all outcomes is given by the probability vector $\boldsymbol{\theta} = (\theta_{1}, \dots, \theta_{K})$.  This means that the probability $P(x=k)=\theta_{k}$ has support $x \in \{1, \dots, K\}$. The \index{Categorical probability distribution}Categorical probability distribution, denoted $Cat(\boldsymbol{\theta})$, is a generalization of the \index{Bernoulli probability distribution}Bernoulli probability distribution and is sometimes referred to it by its alternative names, i.e. the generalised \index{Bernoulli probability distribution}Bernoulli probability distribution or the Multinoulli probability distribution. In probability theory, the \index{Categorical probability distribution}Categorical probability distribution is often used to explain the outcome of a single experiment with more than two possible outcomes, such as rolling a six-sided die~\cite{ref:wackerly:2014}. The \acs{PMF} of the \index{Categorical probability distribution}Categorical probability distribution is given as

\begin{equation}
      \label{eq:probability:probability_distributions:categorical:pmf}
      P(x \vert \boldsymbol{\theta}; K) = f_{Cat}(x; K, \boldsymbol{\theta}) = \prod_{k=1}^{K}\theta_{k}^{[x = k]}
\end{equation}

where the term, $[x = k]$, is the \textit{Iversion Bracket}~\cite{ref:iverson:1962}, yielding 1 if $x = k$ and 0 otherwise. Given class $k$, the categorical distribution simply yields $\theta_{k}$ as follows:

\begin{equation}
      \label{eq:probability:probability_distributions:categorical:pmf_k}
      f_{Cat}(x=k; K, \boldsymbol{\theta}) = \theta_{k}
\end{equation}

The random variable $x$ can also be encoded in binary format, yielding a vector $\boldsymbol{x} = (x_{1}, \dots, x_{K})$ of \index{Bernoulli probability distribution}Bernoulli probability distributions such that the support is $\forall_{k=1}^{K} x_{k} \in \{0, 1\}$. Importantly, if the outcome of the random event is of category $k$, then $x_{k} = 1$ and $\forall_{j=1}^{K} x_{j} = 0, j \neq k$ so that the standard $K-1$ probability simplex $S$ still holds. The \acs{PMF} of the \index{Categorical probability distribution}Categorical probability distribution is rewritten as

\begin{equation}
      \label{eq:probability:probability_distributions:categorical:pmf_alt}
      f_{Cat}(\boldsymbol{x}; K, \boldsymbol{\theta}) = \prod_{k=1}^{K}\theta_{k}^{\mathbbm{1}_{1}(x_{k})}
\end{equation}

where $\mathbbm{1}(x_{k})$ is the \textit{indicator function}, yielding 1 if $x_{k} = 1$ and 0 otherwise.

Since there is no order to the underlying categories, the mean of the distribution does not yield any relevant information. The mode of the distribution is given as

\begin{equation}
      \label{eq:probability:probability_distributions:categorical:mode}
      M[x] = \argmax_{k}(\theta_{1}, \dots, \theta_{K})
\end{equation}

\subsection{Multinomial Probability Distribution}
\label{sec:probability:probability_distributions:multinomial}

The \index{Multinomial probability distribution}Multinomial probability distribution is a discrete probability distribution over some random variable $\boldsymbol{x} = (x_{1}, \dots\, x_{K})$ that takes on the counts for each occurrence of $K$ possible classes in $N$ independent trials~\cite{ref:wackerly:2014}. The probabilities for all possible outcomes in a single trial is given by the probability vector, $\boldsymbol{\theta} = (\theta_{1}, \dots, \theta_{K})$. The \index{Multinomial probability distribution}Multinomial probability distribution, denoted $Mul(N, K, \boldsymbol{\theta})$, is thus a generalization of the \index{Binomial probability distribution}Binomial probability distribution to $K$ dimensions. Consider the following special cases:

\begin{itemize}
      \item When $K$ is 2 and $N = 1$, the \index{Multinomial probability distribution}Multinomial probability distribution is the \index{Bernoulli probability distribution}Bernoulli probability distribution.
      \item When $K$ is 2 and $N > 1$, the \index{Multinomial probability distribution}Multinomial probability distribution is the \index{Binomial probability distribution}Binomial probability distribution.
      \item When $K > 2$ and $N = 1$, the \index{Multinomial probability distribution}Multinomial probability distribution is the \index{Categorical probability distribution}Categorical probability distribution.
\end{itemize}

The support for the Multinomial is $\forall_{i=1}^{K} x_{k} \in \{1, \dots, N\}, \sum_{k=1}^{K}x_{k} = N$ and the \acs{PMF} for the \index{Multinomial probability distribution}Multinomial probability distribution is given as

\begin{equation}
      \label{eq:probability:probability_distributions:multinomial:pmf}
      P(\boldsymbol{x} \vert \boldsymbol{\theta}; N; K) = f_{Mul}(\boldsymbol{x}; N, K, \boldsymbol{\theta}) = \frac{N!}{\prod_{k=1}^{K}x_{k}!} \prod_{k=1}^{K}\theta_{k}^{x_{k}}
\end{equation}

Similar to the \index{Categorical probability distribution}Categorical probability distribution, the random variable $x$ can also be encoded in binary format, yielding an $N \times K$ matrix $\boldsymbol{X}$ of \index{Bernoulli probability distribution}Bernoulli probability distributions. The support is then given as $\boldsymbol{X} \in \{0, 1\}^{N \times K}, \forall_{i=1}^{N}\sum_{k=1}^{K} x_{i,k} = 1$ so that the standard $K-1$ probability simplex $S$ still holds for each trial. The \acs{PMF} of the \index{Multinomial probability distribution}Multinomial probability distribution is then rewritten as

\begin{equation}
      \label{eq:probability:probability_distributions:multinomial:pmf_alt}
      \begin{split}
            f_{Mul}(\boldsymbol{X}; N, K, \boldsymbol{\theta}) &= \frac{N!}{\prod_{k=1}(\sum_{i=1}^{N}x_{i, k})!}\prod_{i=1}^{N}\prod_{k=1}^{K}\theta_{k}^{\mathbbm{1}_{1}(x_{i,k})} \\
            &= \frac{N!}{\prod_{k=1}(\sum_{i=1}^{N}x_{i, k})!} \prod_{k=1}^{K}\theta_{k}^{\sum_{i=1}^{N}\mathbbm{1}_{1}(x_{i,k})} \\
            &= \frac{N!}{\prod_{k=1}(\sum_{i=1}^{N}x_{i, k})!} \prod_{k=1}^{K}\theta_{k}^{N_{k}} \\
      \end{split}
\end{equation}

where $N_{k}$ is a summary variable, denoting the number of times a category $k$ occurs over all trials in $N$.

@@

\section{Summary}\label{sec:probability:summary}

This chapter provided all the necessary background information on probability theory and statistics. The origins of statistical analysis were discussed and the differences between the frequentist and Bayesian view of statistics were discussed in detail. Probability distributions and proofs of conjugate priors were provided with detailed mathematical descriptions. Finally, a detailed description of Bayesian analysis was provided with detailed mathematical descriptions.

This chapter concludes all the relevant background information that is needed to formulate the \acs{BHH}. The proposed \acs{BHH} is presented in the following chapter.

>>>>

\section{Probability}
\label{sec:probability}

@@

Chapter \ref{chap:hhs} provided the concept of a \acf{HH}. This dissertation aims to develop a selection \acs{HH} that makes use of probability theory to select the best \acs{HH} to solve a given problem. This chapter aims to provide the necessary background information on probability theory and statistics. These are large fields and focus is put on the elements that are required to formulate the proposed \acf{BHH}.

@@

@@

@@
